---
title: "10 - Tibbles"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 10.2 Creating Tibbles

Convert regular `data.frame` to a `tibble` and create a new `tibble`.

```{r}
as_tibble(iris)

tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

# 10.3 Tibble vs data.frame

The two main differences are printing and subsetting.

## 10.3.1 Priting

By default it only prints the first 10 rows but we can control this using `print()`

```{r, eval = FALSE}
nycflights13::flights %>% 
  print(n = 3, width = Inf)
```

## 10.3.2 Subsetting

This is how we can select specific columns:

- `df[[ name or posision ]]``
- `df$name`

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

df$x
df[["x"]]
df[[1]]
```

We can also use this in a pipe using the `.` placeholder.

```{r}
df %>% .$x
df %>% .[["x"]]
```

# 10.4 Interacting with older code

We can also return to the classic R `data.frame`

```{r}
class(as.data.frame(df))

as.data.frame(df)
```

# 10.5  Exercises

1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).

`typeof()` doesn't work, as it says both are a `list`.

```{r}
typeof(mtcars)
typeof(diamonds)
```

Use `is_tibble()`.

```{r}
is_tibble(mtcars)
is_tibble(diamonds)
```

Printing the objects is different between the two 

2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

```{r}
df <- data.frame(abc = 1, xyz = "a")
df
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

```{r}
tbl <- tibble(
  abc = 1, 
  xyz = "a"
)
tbl
tbl$x
tbl[, "xyz"]
tbl[, c("abc", "xyz")]
```

Its not really expected that `df$x` to work as the column name is `xyz` not just `x`, `data.frame` will complete but its not consistent as changing the data might make it complete differently without changing the code.

Also `data.frame` returns differint types if you index one columns (vector) or index multiple columns (another data.frame). A tibble always returns another tibble when indexing.

3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

We can just use the `[]` and `[[]]` expressions. The first one returns a tibble and the second one the vector of that columns.

```{r}
var <- "mpg"

cars_tbl <- as_tibble(mtcars)
cars_tbl[var]
cars_tbl[[var]]
```

4. Practice referring to non-syntactic names in the following data frame by:

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

annoying
```

- Extracting the variable called 1.

```{r}
# annoying$1 # Doesn't work
annoying$`1`
annoying["1"]
```

- Plotting a scatterplot of 1 vs 2.

```{r}
annoying %>% 
  ggplot() + 
  geom_point(aes(x = `1`, y = `2`))
```

- Creating a new column called 3 which is 2 divided by 1.

```{r}
annoying <-
  annoying %>%
  mutate(`3` = `2` / `1`)
annoying
```

- Renaming the columns to one, two and three.

```{r}
annoying %>%
  rename(one = `1`, two = `2`, three = `3`)
```

5. What does tibble::enframe() do? When might you use it?

```{r}
?enframe
```

It converts named vectors or list to two-column data frames.

```{r}
enframe(c(a = 1, b = 2, c = 3))
enframe(list(female = 1, male = 2))
```

6. What option controls how many additional column names are printed at the footer of a tibble?

`options(tibble.width = Inf)` for all columns to be printed.
