---
title: "11 - Data Import"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 11.2 Getting started

```{r}
heights <- read_csv("data/heights.csv")
```

Some common arguments:

```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)
read_csv("1,2,3\n4,5,6", col_names = FALSE)
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
read_csv("a,b,c\n1,2,.", na = ".")
```

## 11.2.2 Exercises

1. What function would you use to read a file where fields were separated with “|”?

`read_delim()`

```{r}
read_delim("a|b|c\n1|2|3", delim = "|")
```

2. Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?

```{r}
?read_csv
?read_tsv
```

A lot (all?) they both use `read_delim` with different defaults. For example: `col_names`, `quote`, `na`, `n_max`.

3. What are the most important arguments to read_fwf()?

```{r}
?read_fwf
```

`col_positions`, becuase thats how it knows where to look for values on the columns.

4. Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you’ll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame?

```
"x,y\n1,'a,b'"
```

```{r}
read_csv("x,y\n1,'a,b'", quote="'")
read_delim("x,y\n1,'a,b'", delim = ",",  quote = "'")
```

5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code?

```{r}
read_csv("a,b\n1,2,3\n4,5,6")
```

There is only two columns in the first row so `read_csv()` will read those two columns and ignore the other values.

```{r}
read_csv("a,b,c\n1,2\n1,2,3,4")
```

There is 3 cols in the header, 2 values in one row, 4 values in the second row. The resulting tibble is of 3 columns, first one has an extra missing value and the 4th column values are ignored.

```{r}
read_csv("a,b\n\"1")
```

Missing value for the `b` column is added for the first row that doesn't have that value.

```{r}
read_csv("a,b\n1,2\na,b")
```

It reads the values fine but both columns are parsed as character columns.

```{r}
read_csv("a;b\n1;3")
```

The delimiter is not set correctly to `;`, it should be:

```{r}
read_delim("a;b\n1;3", delim = ";")
```

# 11.3 Parsing a vector

Take a character vector and return an specialised vector.

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
str(parse_integer(c("1", "2", "3")))
str(parse_date(c("2010-01-01", "1979-10-14")))
```

Failures will be saved in a tibble

```{r}
x <- parse_integer(c("123", "345", "abc", "123.45"))
x
problems(x)
```

## 11.3.1 Numbers

`readr` has the concept of `locale()` to handle different formats.

```{r}
parse_double("1.23")
parse_double("1,23", locale = locale(decimal_mark = ","))
```

```{r}
parse_number("$100")
#> [1] 100
parse_number("20%")
parse_number("It cost $123.45")
```

```{r}
parse_number("$123,456,789")
parse_number("123.456.789", locale = locale(grouping_mark = "."))
parse_number("123'456'789", locale = locale(grouping_mark = "'"))
```

## 11.3.2 Strings

`readr` uses `UTF-8` as default for reads and writes!

We can use `locale()` also for strings.

```{r}
x1 <- "El Ni\xf1o was particularly bad this year"
x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"
x1
x2
```

```{r}
parse_character(x1, locale = locale(encoding = "Latin1"))
parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

```{r}
guess_encoding(charToRaw(x1))
guess_encoding(charToRaw(x2))
```

## 11.3.3 Factors

Similar to parsing numbers, error will be put in a dataframe.

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

## 11.3.4 Dates, date-times, and times

- date (the number of days since 1970-01-01)
- date-time (the number of seconds since midnight 1970-01-01)
- time (the number of seconds since midnight)

`parse_datetime()` (by default) expects an ISO8601 date.

```{r}
parse_datetime("2010-10-01T2010")
parse_datetime("20101010")
```

```{r}
parse_date("2010-10-01")
```

```{r}
library(hms)
parse_time("01:10 am")
parse_time("20:10:01")
```

We can always overwrite with `format`.

```{r}
parse_date("01/02/15", format = "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")
```

## 11.3.5 Exercises

1. What are the most important arguments to locale()?

```{r}
?locale
```

`encoding` is probably my number 1. After that `date_format`, `time_format` and `tz`.

2. What happens if you try and set decimal_mark and grouping_mark to the same character? 

Cannot be the same:

```{r}
parse_number("123.32", locale = locale(decimal_mark = ".", grouping_mark = "."))
```

- What happens to the default value of grouping_mark when you set decimal_mark to “,”?
- What happens to the default value of decimal_mark when you set the grouping_mark to “.”?

The defaults are inverted to the value that is set.

```{r}
locale(decimal_mark = ",")
locale(grouping_mark = ",")
```

3. I didn’t discuss the date_format and time_format options to locale(). What do they do? Construct an example that shows when they might be useful.

They are the formats for time and dates:

```{r}
parse_date("05-02-00", locale = locale(date_format = "%d-%m-%y"))
parse_time("02-00-08 am", locale = locale(time_format = "%M-%S-%I %p"))
```

4. If you live outside the US, create a new locale object that encapsulates the settings for the types of file you read most commonly.

```{r}
locale(date_names = "es",
       date_format = "%Y/%m/%d",
       time_format = "%H/%M/%S",
       decimal_mark = ".")
```

5. What’s the difference between read_csv() and read_csv2()?

```{r}
?read_csv
?read_csv2
```

`read_csv()` uses `,` as delimiter. `read_csv2()` uses `;` as delimiter.

6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.

7. Generate the correct format string to parse each of the following dates and times:

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
```

```{r}
parse_date(d1, locale = locale(date_format = "%B %d, %Y"))
parse_date(d2, locale = locale(date_format = "%Y-%b-%d"))
parse_date(d3, locale = locale(date_format = "%d-%b-%Y"))
parse_date(d4, "%B %d (%Y)")
parse_date(d5, "%m/%d/%y")
parse_time(t1, "%H%M")
parse_time(t2, "%I:%M:%OS %p")  # OS: real seconds
```

# 11.4 Parsing a file

`readr` uses some conservative heuristics to guess the types of columns by reading the first 1000 rows. We can use `guess_parser()` to emulate this.

```{r}
guess_parser("2010-10-01")
guess_parser(c("1", "5", "9"))

str(parse_guess("2010-10-10"))
```

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
problems(challenge)
```

We can always overwrite the heuristics.

```{r}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
tail(challenge)
```

# 11.5 Writing to a file

`write_csv()` and `write_tsv()` can be used to save files to disk. The default is to use `UTF-8` and `ISO8601`.

```{r}
write_csv(challenge, "output/challenge.csv")
```

We can also save this in a R binary format that saves the types and more information

```{r}
write_rds(challenge, "output/challenge.rds")
read_rds("output/challenge.rds")
```

Feather is another alternative that is not specific to R

```{r}
# install.packages("feather")
library(feather)
write_feather(challenge, "output/challenge.feather")
read_feather("output/challenge.feather")
```
