---
title: "12 - Tidy"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 12.2 Tidy data

```{r}
table1
table2
table3
table4a
table4b
```

There is 3 rules for tidy data:

1. Each variable must hve its own column
2. Each observation must have its own row
3. Each value must have its own cell

In the examples above, `table1` is tidy, that makes things easy.

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)

ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

## 12.2.1 Exercises

1. Using prose, describe how the variables and observations are organised in each of the sample tables.

- `table1` is tidy
- `table2` combines the two types (cases and population) in the same column and it combines its values in the `count` column
- `table3` mixes the cases and population into the ratio column, that column also has two values per cell
- `table4a` and `table4b` split the data into two datasets one for cases and one for population

2. Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

- Extract the number of TB cases per country per year.
- Extract the matching population per country per year.
- Divide cases by population, and multiply by 10000.
- Store back in the appropriate place.

```{r}
table2

t2_cases <- table2 %>% 
  filter(type == "cases")

t2_pop <- table2 %>%
  filter(type == "population")

t2_cases_per_cap <- bind_cols(t2_cases, t2_pop) %>% 
  mutate(count = 10000 * count / count1) %>%  # We name it count to its the same format as table2
  mutate(type = "rate") %>% # This is for the join below to work
  select(country, year, type, count)

bind_rows(table2, t2_cases_per_cap) %>%
  arrange(country, year, type, count)
```

```{r}
table4a
table4b

tibble(country = table4a$country,
       `1999` = table4a[["1999"]] / table4b[["1999"]] * 10000,
       `2000` = table4a[["2000"]] / table4b[["2000"]] * 10000)
```

- Which representation is easiest to work with? Which is hardest? Why?

In a way it's easier to do it with table4a and table4b since they are more tidy even if the columns are weird and its two datasets.

3. Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?

First we need to filter by cases, after that it's fine.

```{r}
table2 %>% 
  filter(type == "cases") %>% 
  ggplot(aes(year, count), colour = "grey50") + 
    geom_line(aes(group = country)) +
    geom_point(aes(colour = country))
```

# 12.3 Spreading and gathering

There is usually two common problem:

1. One variable might be spread across multiple columns. `gather()`.
2. One observation might be scattered across multiple rows. `spread()`.

## 12.3.1 Gathering

This is when the columns names represent a value.

```{r}
table4a
table4b
```

We can solve it with `gather()` and `left_join()`

```{r}
tidy4a <- table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")

tidy4b <- table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")

left_join(tidy4a, tidy4b)
```

## 12.3.2 Spreading

This is the opposite of gathering. Its when one observation is scattered across multiple rows.

```{r}
table2
```

We can solve it with `spread()`

```{r}
table2 %>% 
  spread(key = type, value = count)
```

## 12.3.3 Exercises

1. Why are gather() and spread() not perfectly symmetrical?
Carefully consider the following example:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half   = c(   1,   2,    1,   2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```

(Hint: look at the variable types and think about column names.)

Year gets converted to a `chr` column.

Both spread() and gather() have a convert argument. What does it do?

That argument will try to convert the `key` columns to it's correct class.

```{r}
stocks %>% 
  spread(year, return, convert = T) %>% 
  gather("year", "return", `2015`:`2016`, convert = T)
```

2. Why does this code fail?

```{r}
table4a %>% 
  gather(1999, 2000, key = "year", value = "cases")
#> Error in inds_combine(.vars, ind_list): Position must be between 0 and n
```

Because its using a number as the column name, if the column name is a number it has to use "`".

3. Why does spreading this tibble fail? How could you add a new column to fix the problem?

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```

```{r}
people %>% 
  spread(key = key, value = value)
```

It fail because there is a duplicated key for "Pillip Wood" + "age", it probably means that it was taken at two points in time, we should add that column.

```{r}
people %>%
  mutate(time = c(1, 2, 2, 1, 1)) %>%
  select(time, everything()) %>%
  spread(key, value)
```

4. Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
preg
```

We need to gather it so the gender value is placed in a column.

```{r}
preg %>% 
  gather(male, female, key = "gender", value = "value")
```

# 12.4 Separating and uniting

How to fix cells having multiple values

```{r}
table3
```

## 12.4.1 `separate()`

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/", convert = T)
```

## 12.4.2 `unite()`

```{r}
table5
```

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

## 12.4.3 Exercises

1. What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
```

```{r}
?separate
```

`fill`: If sep is a character vector, this controls what happens when there are not enough pieces. There are three valid options:
- "warn" (the default): emit a warning and fill from the right
- "right": fill with missing values on the right
- "left": fill with missing values on the left

`extra`: If sep is a character vector, this controls what happens when there are too many pieces. There are three valid options:
"warn" (the default): emit a warning and drop extra values.
"drop": drop any extra values without a warning.
"merge": only splits at most length(into) times

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), sep = ",", extra = "merge")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), sep = ",", fill = "left")
```

2. Both unite() and separate() have a remove argument. What does it do? Why would you set it to FALSE?

```{r}
?unite
?separate
```

`remove`: If TRUE, remove input column from output data frame.

I would set it to false if we want to generate a new variable but keep the old ones to verify and then remove to make the data set tidy.

3. Compare and contrast separate() and extract(). Why are there three variations of separation (by position, by separator, and with groups), but only one unite?

```{r}
?separate
?extract
```

I think both do similar things and one might just be a subset of the other. `extract()` get a regex while `separate()` can receive a regex but can also receive a number to split by position.

There is only one `unite()` because its a more univeral operation, split can be done in many different ways.

# 12.5 Missing values

There is two types of missing values: explict and implicit.

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks
```

- The return for the fourth quarter of 2015 is explicitly missing, because the cell where its value should be instead contains NA.
- The return for the first quarter of 2016 is implicitly missing, because it simply does not appear in the dataset.

```{r}
stocks %>% 
  spread(year, return)
```

Find implicit missing values using `complete()`:

```{r}
stocks %>% 
  complete(year, qtr)
```

Sometimes a missing value means that the previous value should be carrry over:

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment
```

```{r}
treatment %>% 
  fill(person)
```

## 12.5.1 Exercises

1. Compare and contrast the fill arguments to spread() and complete().

```{r}
?spread
?complete
```

With `complete()` you can pass a named list so it fills values differently for each column. With `spread()` its always `NA`.

2. What does the direction argument to fill() do?

```{r}
?fill
```

`direction`: Direction in which to fill missing values. Currently either "down" (the default) or "up".

```{r}
treatment %>% 
  fill(person, .direction = "up")
```

# 12.6 Case Study








