---
title: "15-factors"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(forcats)
```

## 15.2 Creating factors

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
sort(x1)
```

```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
month_levels
```

```{r}
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```

We can always see the `levels`

```{r}
levels(y1)
```


# 15.3 General Social Survey

```{r}
gss_cat
```

We can see there is some variables in the tibble that are factors, such as `race`.

```{r}
gss_cat %>%
  count(race)
```

## 15.3.1 Exercise

1. Explore the distribution of rincome (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

```{r}
gss_cat %>%
  ggplot(aes(x = rincome)) + 
    geom_bar()
```

The labels are impossible to read like that, also the order is not ideal:

```{r}
levels(gss_cat$rincome)
```

```{r}
ordered_levels <- rev(levels(gss_cat$rincome))
ordered_levels
```


```{r}
gss_cat %>%
  mutate(rincome = fct_relevel(rincome, ordered_levels)) %>%
  mutate(rincome = fct_relevel(rincome, c("No answer", "Don't know", "Refused", "Not applicable"))) %>% 
  ggplot(aes(x = rincome)) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 90))
```

2. What is the most common relig in this survey? What’s the most common partyid?

```{r}
gss_cat %>%
  count(relig) %>%
  arrange(-n)
```

```{r}
gss_cat %>%
  count(partyid) %>%
  arrange(-n)
```

3. Which relig does denom (denomination) apply to? How can you find out with a table? How can you find out with a visualisation?

```{r}
gss_cat %>%
  count(relig, denom) %>% 
  filter(!denom %in% c("No answer", "Other", "Don't know", "Not applicable", "No denomination")) %>% 
  arrange(-n)
```

Demon is used for Protestant.

```{r}
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

# 15.4 Modifying factor order

```{r}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()
```

We can use `fct_reorder()`.

```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```

## 15.4.1 Exercises

1. There are some suspiciously high numbers in tvhours. Is the mean a good summary?

```{r}
gss_cat %>%
  ggplot(aes(tvhours)) +
  geom_histogram() +
  geom_vline(xintercept = mean(gss_cat$tvhours, na.rm = TRUE), colour = "red") +
  geom_vline(xintercept = median(gss_cat$tvhours, na.rm = TRUE), colour = "blue")
```

Median is usually always better.

2. For each factor in gss_cat identify whether the order of the levels is arbitrary or principled.

```{r}
fct_gss <- gss_cat[sapply(gss_cat, is.factor)]
lapply(fct_gss, levels)
```

3. Why did moving “Not applicable” to the front of the levels move it to the bottom of the plot?

```{r}
gss_cat %>%
  mutate(rincome = rincome %>% fct_relevel("Not applicable")) %>%
  ggplot(aes(rincome)) +
  geom_bar()

gss_cat %>%
  mutate(rincome = rincome %>% fct_relevel("Not applicable")) %>%
  ggplot(aes(rincome)) +
  geom_bar() +
  coord_flip()
```

It keeps the order but looks like it didn't.

# 15.5 Modifying factor levels

This allows you to clarify labels for publication, and collapse levels for high-level displays.

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)
```

We can also collapse categories using `fct_collapse()`.

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

We can use `fct_lump()` to collapse the groups with less items.


## 15.5.1 Exercises

1. How have the proportions of people identifying as Democrat, Republican, and Independent changed over time?

```{r}
gss_cat %>%
  mutate(partyid =
           fct_collapse(partyid,
                        other = c("No answer", "Don't know", "Other party"),
                        rep = c("Strong republican", "Not str republican"),
                        ind = c("Ind,near rep", "Independent", "Ind,near dem"),
                        dem = c("Not str democrat", "Strong democrat"))) %>%
  count(year, partyid)  %>%
  group_by(year) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(x = year, y = p,
             colour = fct_reorder2(partyid, year, p))) +
    geom_point() +
    geom_line() +
    labs(colour = "Party ID.")
```


2. How could you collapse rincome into a small set of categories?

```{r}
gss_cat %>%
  mutate(rincome =
            fct_collapse(rincome,
              `Unknown` = c("No answer", "Don't know", "Refused", "Not applicable")
            )) %>%
  ggplot(aes(x = rincome)) +
    geom_bar() +
    coord_flip()
```
















