---
title: "18-pipes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(magrittr)
```

Pipes `%>%` come from `magirttr`.

# 18.2 Alternatives

Creating variables

```r
foo_foo_1 <- hop(foo_foo, through = forest)
foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
foo_foo_3 <- bop(foo_foo_2, on = head)
```

This is not terrible for memory since R is smart and will only copy columnn from dataframe when they change:

```{r}
# install.packages("pryr")
```


```{r}
library(tidyverse)
library(pryr)

diamonds <- ggplot2::diamonds
diamonds2 <- diamonds %>% 
  dplyr::mutate(price_per_carat = price / carat)

pryr::object_size(diamonds)
#> 3.46 MB
pryr::object_size(diamonds2)
#> 3.89 MB
pryr::object_size(diamonds, diamonds2)
#> 3.89 MB
```

```{r}
diamonds$carat[1] <- NA
pryr::object_size(diamonds)
#> 3.46 MB
pryr::object_size(diamonds2)
#> 3.89 MB
pryr::object_size(diamonds, diamonds2)
#> 4.32 MB
```

The advantage of pipes is that it focuses on `verbs` not `nouns`.

```{r}
foo_foo %>%
  hop(through = forest) %>%
  scoop(up = field_mice) %>%
  bop(on = head)
```

Internally this gets translated to something like:

```{r}
my_pipe <- function(.) {
  . <- hop(., through = forest)
  . <- scoop(., up = field_mice)
  bop(., on = head)
}
my_pipe(foo_foo)
```

Pipes **won't** work for functions that:

1. Use variables in the environment

```{r}
assign("x", 10)
x

"x" %>% assign(100)
x
```

To use it you have to pass the environment explictly:

```{r}
env <- environment()
"x" %>% assign(100, envir = env)
x
```

2. Functions that use lazy evaluation

```{r}
tryCatch(stop("!"), error = function(e) "An error")
#> [1] "An error"

stop("!") %>% 
  tryCatch(error = function(e) "An error")
```

# 18.3 When not to use pipes

- Pipes are longer than (around) 10 steps. Create inermediate objects with meaningful names.
- We have mutiple inputs and outputs.
- We are starting to think of a graph structure, pipes are fundamentally linear.

# 18.4 Other tools from magrittr

- The "tee" pipe: `%T>%` to return the left-hand side instead of the right hand. In this case its calling `str()` of nothing:

```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()
```

Here it will call `str()` of `matrix()`.

```{r}

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

- Use `%$%` when we want to expand the dataframe columns as variables

```{r}
mtcars %$%
  cor(disp, mpg)
```

- Assign using `%<>%`

```
mtcars <- mtcars %>% 
  transform(cyl = cyl * 2)
```

```{r}
mtcars %<>% transform(cyl = cyl * 2)
```

Hadley doesn't like this one that much, better to be explicit.








