---
title: "19-functions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(lubridate)
library(tidyverse)
```


```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```

Becomes

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

## 19.2.1 Practice

1. Why is TRUE not a parameter to rescale01()? What would happen if x contained a single missing value, and na.rm was FALSE?

```{r}
rescale01_alt <- function(x, na.rm = FALSE) {
  rng <- range(x, na.rm = na.rm)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01_alt(c(NA, 1:5), na.rm = FALSE)
rescale01_alt(c(NA, 1:5), na.rm = TRUE)
```

`range()` will return all `NA`s if there is one `NA` value.

2. In the second variant of rescale01(), infinite values are left unchanged. Rewrite rescale01() so that -Inf is mapped to 0, and Inf is mapped to 1.

```{r}
rescale01_alt2 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  y <- (x - rng[1]) / (rng[2] - rng[1])
  y[y == -Inf] <- 0
  y[y == Inf] <- 1
  y
}

rescale01_alt2(c(Inf, -Inf, 0:5, NA))
```


3. Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need? Can you rewrite it to be more expressive or less duplicative?

```r
mean(is.na(x))
```

```{r}
prop_na <- function(x) {
  mean(is.na(x))
}
prop_na(c(0, 1, 2, NA, 4, NA))
```

```
x / sum(x, na.rm = TRUE)
```

```{r}
prop_mean <- function(x) {
  x / sum(x, na.rm = TRUE)
}

my_mean(c(0, 1, 2, NA, 4, NA))
```

```r
sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
```

```{r}
coef_variation <- function(x, na.rm = FALSE) {
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}
coef_variation(1:5)
coef_variation(c(1:5, NA))
coef_variation(c(1:5, NA), na.rm = TRUE)
```

4. Follow http://nicercode.github.io/intro/writing-functions.html to write your own functions to compute the variance and skew of a numeric vector.

```{r}
variance <- function(x, na.rm = TRUE) {
  n <- length(x)
  m <- mean(x, na.rm = TRUE)
  sq_err <- (x - m) ^ 2
  sum(sq_err) / (n - 1)
}
var(1:10)
variance(1:10)
```

```{r}
skewness <- function(x, na.rm = FALSE) {
  n <- length(x)
  m <- mean(x, na.rm = na.rm)
  s <- sd(x, na.rm = na.rm)
  n * sum(((x - m) / s) ^ 3) / (n - 1) / (n - 2)
}
skewness(c(1, 2, 5, 100))
```

5. Write both_na(), a function that takes two vectors of the same length and returns the number of positions that have an NA in both vectors.

```{r}
both_na <- function(x, y) {
  stopifnot(length(x) == length(y))
  count <- 0
  for (i in seq_along(x)) {
    if (is.na(x[i])  && is.na(y[i])) {
      count <- count + 1
    }
  }
  count
}
both_na(c(NA, NA,  1, 2),
        c(NA,  1, NA, 2))
```

```{r}
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}
both_na(c(NA, NA,  1, 2),
        c(NA,  1, NA, 2))
```

6. What do the following functions do? Why are they useful even though they are so short?

```{r}
is_directory <- function(x) file.info(x)$isdir
is_readable <- function(x) file.access(x, 4) == 0

is_directory("~")
is_readable("~")
```

- `is_directory()` checks if the path is a directory
- `is_readable()` checks if the path is readable

7. Read the complete lyrics to “Little Bunny Foo Foo”. There’s a lot of duplication in this song. Extend the initial piping example to recreate the complete song, and use functions to reduce the duplication.

```{r}
# :)
```

# 19.3 Functions are for humans

## 19.3.1 Exercises

1. Read the source code for each of the following three functions, puzzle out what they do, and then brainstorm better names.

```{r}
f1 <- function(string, prefix) {
  substr(string, 1, nchar(prefix)) == prefix
}
```

`f1()` is like `startswith()` in Python

```{r}
f2 <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}
```

`f2()` is like droping the last element of the list.

```{r}
f3 <- function(x, y) {
  rep(y, length.out = length(x))
}
```

`f3()` is converting the length of vector `y` to `x`.

2. Take a function that you’ve written recently and spend 5 minutes brainstorming a better name for it and its arguments.

```{r}
# :)
```

3. Compare and contrast rnorm() and MASS::mvrnorm(). How could you make them more consistent?

- `rnorm()` samples from the univariate normal distribution
- `MASS::mvrnorm` samples from the multivariate normal distribution

4. Make a case for why norm_r(), norm_d() etc would be better than rnorm(), dnorm(). Make a case for the opposite.

They both do some normalization (i image) so they are part of the same family.
On the other hand if there are different types of operations might be better to do the second option.

# 19.4 Conditional execution

## 19.4.4 Exercises

1. What’s the difference between if and ifelse()? Carefully read the help and construct three examples that illustrate the key differences.

`ifelse()` tests the conditions in vectors and return a vector. `if` is a single condition.

2. Write a greeting function that says “good morning”, “good afternoon”, or “good evening”, depending on the time of day. (Hint: use a time argument that defaults to lubridate::now(). That will make it easier to test your function.)

```{r}
greet <- function(time = lubridate::now()) {
  hr <- lubridate::hour(time)
  # I don't know what to do about times after midnight,
  # are they evening or morning?
  if (hr < 12) {
    print("good morning")
  } else if (hr < 17) {
    print("good afternoon")
  } else {
    print("good evening")
  }
}

greet()
greet(ymd_h("2017-01-08:05"))
greet(ymd_h("2017-01-08:13"))
greet(ymd_h("2017-01-08:20"))
```

3. Implement a fizzbuzz function. It takes a single number as input. If the number is divisible by three, it returns “fizz”. If it’s divisible by five it returns “buzz”. If it’s divisible by three and five, it returns “fizzbuzz”. Otherwise, it returns the number. Make sure you first write working code before you create the function.

```{r}
fizzbuzz <- function(x) {
  stopifnot(is.numeric(x))
  if (!(x %% 3) && !(x %% 5)) {
    "fizzbuzz"
  } else if (!(x %% 3)) {
    "fizz"
  } else if (!(x %% 5)) {
    "buzz"
  } else {
    x
  }
}

sapply(1:20, fizzbuzz)
```

4. How could you use cut() to simplify this set of nested if-else statements?

```
if (temp <= 0) {
  "freezing"
} else if (temp <= 10) {
  "cold"
} else if (temp <= 20) {
  "cool"
} else if (temp <= 30) {
  "warm"
} else {
  "hot"
}
```

```{r}
temp <- seq(-10, 50, by = 5)
cut(temp, c(-Inf, 0, 10, 20, 30, Inf), right = TRUE, labels = c("freezing", "cold", "cool", "warm", "hot"))
```

How would you change the call to cut() if I’d used < instead of <=? What is the other chief advantage of cut() for this problem? (Hint: what happens if you have many values in temp?)

```{r}
temp <- seq(-10, 50, by = 5)
cut(temp, c(-Inf, 0, 10, 20, 30, Inf), right = FALSE, labels = c("freezing", "cold", "cool", "warm", "hot"))
```

5. What happens if you use switch() with numeric values?

```{r}
?switch
```

If `n` is numeric, it will return the `n`th argument from `...`.
This means that if `n = 1`, `switch()` will return the first argument in `...`,
if `n = 2`, the second, and so on.

```{r}
switch(1, 4,5,6)
switch(2, 4,5,6)
```

6. What does this switch() call do? What happens if x is “e”? Experiment, then carefully read the documentation.

```{r}
switch("e", 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)

```

# 19.5 Function arguments

Two types of arguments:

1. data
2. details (usually optional with good defaults)

## 19.5.3 Checking values (stop)

We can use `stop()` and `stopif()`.

```{r}
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length", call. = FALSE)
  }
  sum(w * x) / sum(w)
}
wt_mean(1:6, 2:6)
```


```{r}
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}
wt_mean(1:6, 6:1, na.rm = "foo")
```

## 19.5.3 Dot-dot-dot (…)

This is like `*args` on Python.

```{r}
commas <- function(...) {
  str_c(..., collapse = ", ")
}

commas(letters[1:10])
```


## 19.5.5 Exercises

1. What does `commas(letters, collapse = "-")` do? Why?

```{r error=TRUE}
commas(letters, collapse = "-")
```

Doesn't work since the collapse argument is send using  `...`. We can fix this by defining it as an actual argument.

```{r}
commas <- function(..., collapse = ", ") {
  stringr::str_c(..., collapse = collapse)
}

commas(letters, collapse = "-")
```

2. It’d be nice if you could supply multiple characters to the pad argument, e.g. `rule("Title", pad = "-+")`. Why doesn’t this currently work? How could you fix it?

```{r}
rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}

rule("Important output")
rule("Important output", pad = "-+")
```

```{r}
rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  pad_char <- nchar(pad)
  cat(title, " ", stringr::str_dup(pad, width / pad_char), "\n", sep = "")
}
rule("my title", pad = "-+")
```

3. What does the trim argument to mean() do? When might you use it?

Removes the `X%` items of boths sides of the vector. Can be used to remove outliers if the vector is in order.

```{r}
mean(c(99, 1:10))
mean(c(99, 1:10), trim = 0.1)
```

4. The default value for the method argument to cor() is c("pearson", "kendall", "spearman"). What does that mean? What value is used by default?

```{r}
?cor
```

The first value is used by default and you have the option of specifying any of the three values.

# 19.6 Return values

## 19.6.1 Explicit return statements

```{r}
complicated_function <- function(x, y, z) {
  if (length(x) == 0 || length(y) == 0) {
    return(0)
  }
    
  # Complicated code here
}
```

## 19.6.2 Writing pipeable functions

There is basically two types:

1. With transformations, an object is passed to the function’s first argument and a modified object is returned. 2. With side-effects, the passed object is not transformed. Instead, the function performs an action on the object, like drawing a plot or saving a file. Side-effects functions should “invisibly” return the first argument, so that while they’re not printed they can still be used in a pipeline.


```{r}
show_missings <- function(df) {
  n <- sum(is.na(df))
  cat("Missing values: ", n, "\n", sep = "")
  
  invisible(df)
}

# show_missings(mtcars)

mtcars %>% show_missings()
```








































