---
title: "20-vectors"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 20.2 Vector basics

There is two types:

1. Atomic vectors, of which there are six types: logical, integer, double, character, complex, and raw. Integer and double vectors are collectively known as numeric vectors.
2. Lists, which are sometimes called recursive vectors because lists can contain other lists.

Atomic vectors are homogeneous, while lists can be heterogeneous. 

```{r}
typeof(letters)
typeof(1:10)
typeof(list("a", 1, T))
```

```{r}
length(letters)
length(1:10)
length(list("a", 1, T))
```

# 20.3 Important types of atomic vector

## 20.3.5 Exercises

1. Describe the difference between is.finite(x) and !is.infinite(x).

```{r}
x <- c(0, NA, NaN, Inf, -Inf)
is.finite(x)
is.infinite(x)
```

The main difference is in `NA` and `NaN` as neither function considers them to be a part of their set.

2. Read the source code for dplyr::near() (Hint: to see the source code, drop the ()). How does it work?

```{r}
dplyr::near
```

```{r}
.Machine$double.eps^0.5
```

`near()` works by calculating the absolute value of the difference and comparing that to a toleration value.
By default the tolerance is set to the square root of `.Machine$double.eps`, which is the smallest floating point number that the computer can represent.

3. A logical vector can take 3 possible values. How many possible values can an integer vector take? How many possible values can a double take? Use google to do some research.

The number of integers depends on the processor (and R) used, for 32 bit will be able to handle:

```{r}
2^(32-1)  # minus one for the sign
.Machine$integer.max  # One less to represent NA_integer_
```

For doubles R uses 64 bit and conforms with the IEC 60559:

```{r}
2e-308
.Machine$double.xmax
```

4. Brainstorm at least four functions that allow you to convert a double to an integer. How do they differ? Be precise.

```{r}
x <- seq(-10, 10, by = 0.5)
x
round(x)
floor(x)
ceiling(x)
```

5. What functions from the readr package allow you to turn a string into logical, integer, and double vector?

```{r}
parse_logical(c("TRUE", "FALSE", "1", "0", "true", "t", "NA"))
parse_integer(c("1235", "0134", "NA"))
parse_number(c("1.0", "3.5", "$1,000.00", "NA"))
```

# 20.4 Using atomic vectors

## 20.4.1 Coercion

Two types:

1. Explicit: using as.logical(), as.integer(), as.double(), or as.character()
2. Implicit when using a vector type in an operation or context that expects another type, e.g. logical + numeric

For example implicit:

```{r}
x <- sample(20, 100, replace = TRUE)
y <- x > 10
sum(y)  # how many are greater than 10?
mean(y) # what proportion are greater than 10?(y)
```

When creating a vector with multiple types, the most complex always wins:

```{r}
typeof(c(TRUE, 1L))
typeof(c(1L, 1.5))
typeof(c(1.5, "a"))
```

## 20.4.3 Scalars and recycling rules

In R there is no scalars, a single number is a vector of length 1. That makes it easy to understand that all functions are vectorized. This also means that there is no need to iterate for simple mathematical computations.

```{r}
sample(10) + 100
```

Recycling means that for operations where the vectors have not the same lenght R will expand the short one to the size of the long one:

```{r}
1:10
c(1:3, 1:3, 1:3, 1)
1:10 + 1:3
```

The tidyverse doesnt approve recycling that much since it can hide error, it has to be done manually with `rep()`

```{r}
tibble(x = 1:4, y = 1:2)
```

```{r}
tibble(x = 1:4, y = rep(1:2, 2))
```

## 20.4.4 Naming vectors

All types of vectors can be named. You can name them during creation with c():

```{r}
c(x = 1, y = 2, z = 4)
```

Or after the fact with purrr::set_names():

```{r}
set_names(1:3, c("a", "b", "c"))
```

# 20.4.5 Subsetting

For tibbles we used `filter()` for vectors we use `[]`, R is 1-index.

```{r}
x <- c("one", "two", "three", "four", "five")
x[c(3, 2, 5)]

# Negative values drop the elements at the specified positions:
x[c(-1, -3, -5)] 

# It’s an error to mix positive and negative values:
x[c(1, -1)]
```

Subsetting with logic vectors keep `TRUE`s of the operation.

```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)

# All non-missing values of x
x[!is.na(x)]

# All even (or missing!) values of x
x[x %% 2 == 0]
```

Named vectors can be subsetted with char vectors:

```{r}
x <- c(abc = 1, def = 2, xyz = 5)
x[c("xyz", "def")]
```

## 20.4.6 Exercises

1. What does mean(is.na(x)) tell you about a vector x? What about sum(!is.finite(x))?

`mean(is.na(x))` is the proportion of values that are NAs.
`sum(!is.finite(x))` is the number of values that are `NA`, `NaN` or infinites

2. Carefully read the documentation of is.vector(). What does it actually test for? Why does is.atomic() not agree with the definition of atomic vectors above?

```{r}
?is.vector
```

From docs: is.vector returns TRUE if x is a vector of the specified mode having no attributes other than names. It returns FALSE otherwise.

```{r}
?is.atomic
```

is.atomic returns TRUE if x is of an atomic type (or NULL) and FALSE otherwise.

3. Compare and contrast setNames() with purrr::set_names().

In general `set_names()` is more strict as the length of the vectors need to match.

`set_names()` allows for using a function or formula to transform the existing names.

```{r}
purrr::set_names(c(a = 1, b = 2, c = 3), toupper)
```

4. Create functions that take a vector as input and returns:

- The last value. Should you use [ or [[?

```{r}
last_value <- function(x) {
  if (length(x) > 0) {
    return(x[[length(x)]])
  }
  else {
    return(x)
  }
}

last_value(numeric())
last_value(1)
last_value(1:10)
```

- The elements at even numbered positions.

```{r}
even_indices <- function(x) {
  if (length(x) > 0) {
    x[seq_along(x) %% 2 == 0]
  } else {
    x
  }  
}
even_indices(numeric())
even_indices(1)
even_indices(1:10)
```

- Every element except the last value.

```{r}
not_last <- function(x) {
  if (length(x) > 0) {
    x[1:length(x) - 1]
  } else {
    x
  }
}
not_last(1:5)
```

- Only even numbers (and no missing values).

```{r}
even_numbers <- function(x) {
  x[(x %% 2 == 0) & !is.na(x) & !is.nan(x) & !is.infinite(x)]
}
even_numbers(c(0:4, NA, NaN, Inf))
```

5. Why is x[-which(x > 0)] not the same as x[x <= 0]?

```{r}
x <- c(-1:1, Inf, -Inf, NaN, NA)
x[-which(x > 0)]
x[x <= 0]
```

The expressions  `x[-which(x > 0)]` and `x[x <= 0]` return the same values except
for a `NaN` instead of a `NA` in the `which()` based expression.

6. What happens when you subset with a positive integer that’s bigger than the length of the vector? What happens when you subset with a name that doesn’t exist?

```{r}
x <- c(10, 20)
x[3]
x[3:5]
```

It returns a vector of `NA`s
