---
title: "21-iteration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 21.2 For Loops

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

```{r}
output <- vector("double", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output
```

##21.2.1 Exercises

1. Write for loops to, Think about the output, sequence, and body before you start writing the loop:

```{r}
func_apply <- function(df, fun, type) {
  ret = vector(type, length(df))
  for (column in seq_along(df)) {
    ret[column] <- fun(df[[column]])  
  }
  ret
}
```

- Compute the mean of every column in mtcars.

```{r}
func_apply(mtcars, mean, "numeric")
```

- Determine the type of each column in nycflights13::flights.

```{r}
func_apply(mtcars, class, "character")
```

- Compute the number of unique values in each column of iris.

```{r}
unique_vals <- function(x) {
  sum(table(unique(x)))
}
func_apply(iris, unique_vals, "numeric")
```

- Generate 10 random normals for each of u = -10, 0, 10 and 100.

```{r}
ten_draws <- function(x) {
  rnorm(10, mean = x)
}
map(c(-10, 0, 10, 100), ten_draws)
```

2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

```{r}
str_c(letters, collapse = "")
```


```{r}
x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))
sd
```

```{r}
sd(x)
```

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```

```{r}
cumsum(x)
```

3. Combine your function writing and for loop skills:

- Write a for loop that `prints()` the lyrics to the children’s song “Alice the camel”.
- Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any number of people in any sleeping structure.
- Convert the song “99 bottles of beer on the wall” to a function. Generalise to any number of any vessel containing any liquid on any surface.

4. It’s common to see for loops that don’t preallocate the output and instead increase the length of a vector at each step:

```
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```
How does this affect performance? Design and execute an experiment.

# 21.3 For loop variations

## 21.3.1 Modifying an existing object

Note the use of `[[`.

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]])
}

df
```

## 21.3.2 Looping patterns

1. Looping over elements: `for (x in xs)`. Just like in Python.
2. Looping over numeric indices: `for (i in seq_along(xs))`. Like `enumerate()` in Python.
3. Looping over the names: `for (nm in names(xs))`

## 21.3.3 Unknown output lenght

The optimal way to do this is with a `list`. and then combine them in a vector.

```{r}
means <- c(0, 1, 2)

out <- vector("list", length(means))

for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}
str(out)
```

On this case we use `unlist()` to flattern.

```{r}
str(unlist(out))
```

## 21.3.4 Unknown sequence length

This is where `while()` comes in.

```{r}
flip <- function() sample(c("T", "H"), 1)

flips <- 0
nheads <- 0

while (nheads < 3) {
  if (flip() == "H") {
    nheads <- nheads + 1
  } else {
    nheads <- 0
  }
  flips <- flips + 1
}

flips
```

## 21.3.5 Exercises

1. Imagine you have a directory full of CSV files that you want to read in. You have their paths in a vector, files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), and now want to read each one with read_csv(). Write the for loop that will load them into a single data frame.

```{r}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
```

```{r}
dfs <- vector("list", length(files))
for (i in seq_along(files)) {
  dfs[[i]] <- read_csv(files[[i]])
}
df <- bind_rows(df)
df
```

2. What happens if you use for (nm in names(x)) and x has no names? What if only some of the elements are named? What if the names are not unique?

```{r}
x <- 1:3
print(names(x))
```

```{r}
length(NULL)
```

`for (nm in names(x))` wont iterate since the length is zero.

3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example, show_mean(iris) would print:

```{r, eval = FALSE}
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```

```{r}
show_mean <- function(df, digits) {
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      print(sprintf("%s: %.2f", nm, mean(df[[nm]])))
    }
  }
}
show_mean(iris)
```

(Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

```{r}
show_mean <- function(df, digits = 2) {
  maxstr <- max(str_length(names(df)))  # Get max length of all variable names in the dataset
  
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      print(sprintf("%s %.2f",
        str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
        mean(df[[nm]]) ))
    }
  }
}
show_mean(iris)
```

4. What does this code do? How does it work?

```{r}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)

for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}

mtcars
```

There is a list with two "named"" functions that act as transformations.
Then in the for loop for each of the functions in the list, the columns that match a name get a transformation to their values, based on the transform list.

# 21.4 Functional 

```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
col_summary(df, median)
col_summary(df, mean)
```

## 21.4.1 Exercises

1. Read the documentation for `apply()`. In the 2d case, what two for loops does it generalise?

It generalizes looping over the rows or columns of a matrix or data-frame.

2. Adapt `col_summary()` so that it only applies to numeric columns You might want to start with an is_numeric() function that returns a logical vector that has a TRUE corresponding to each numeric column.

```{r}
col_summary2 <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    if (is.numeric(df[[i]])) {
      out[i] <- fun(df[[i]])
    }
  }
  out
}

df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = letters[1:10],
  d = rnorm(10)
)
col_summary2(df, mean)
```

# 21.5 The map functions

The `purrr` package provides:

- map() makes a list.
- map_lgl() makes a logical vector.
- map_int() makes an integer vector.
- map_dbl() makes a double vector.
- map_chr() makes a character vector.

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df
```

```{r}
map_dbl(df, mean)
map_dbl(df, median)
map(df, sd)
```

## 21.5.1 Shortcuts

```{r}
models <- mtcars %>% 
  split(.$cyl) %>% 
  map(function(df) lm(mpg ~ wt, data = df))
models
```

The syntax for creating an anonymous function in R is quite verbose so purrr provides a convenient shortcut: a one-sided formula.

```{r}
models <- mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg ~ wt, data = .))
models
```

# 21.5.3 Exercises

1. Write code that uses one of the map functions to:

- Compute the mean of every column in mtcars.

```{r}
map_dbl(mtcars, mean)
```

- Determine the type of each column in nycflights13::flights.

```{r}
map_chr(nycflights13::flights, typeof)
```

- Compute the number of unique values in each column of iris.

```{r}
map(iris, ~ length(unique(.)))
```

- Generate 10 random normals for each of u = -10, 0, 10, 100

```{r}
map(c(-10, 0, 10, 100), rnorm, n = 10)
```

2. How can you create a single vector that for each column in a data frame indicates whether or not it’s a factor?

```{r}
map_lgl(iris, is.factor)
```

3. What happens when you use the map functions on vectors that aren’t lists? What does map(1:5, runif) do? Why?

```{r}

map(1:5, runif)
```

Map functions work with any vectors, not just lists. As with lists, the map functions will apply the function to each element of the vector.

4. What does map(-2:2, rnorm, n = 5) do? Why? What does map_dbl(-2:2, rnorm, n = 5) do? Why?

This works as expected:

```{r}
map(-2:2, rnorm, n = 5)
```

This fails because `map_dbl` expects the function it applies to each element to return a numeric vector of length one.


```{r}
map_dbl(-2:2, rnorm, n = 5)
```

5. Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function.

```{r}
map(list(mtcars), function(df) lm(mpg ~ wt, data = df))
```


```{r}
map(list(mtcars), ~ lm(mpg ~ wt, data = .))
```

# 21.6 Dealing with failure

`safely(function)` is an adverb that returns a modified function that will not throw an error, instead it will return:

1. restult: if error returns `NULL`
2. error: if no error returns `NULL`

```{r}
safe_log <- safely(log)
str(safe_log(10))
str(safe_log("a"))
```

Purrr provides two other useful adverbs:

`possibly()` always succeeds. It’s simpler than safely(), because you give it a default value to return when there is an error.

```{r}
x <- list(1, 10, "a")
x %>% map_dbl(possibly(log, NA_real_))
```

`quietly()` performs a similar role to safely(), but instead of capturing errors, it captures printed output, messages, and warnings:

```{r}
x <- list(1, -1)
x %>% map(quietly(log)) %>% str()
```

# 21.7 Mapping over multiple arguments

How do we iterate with along multiple inputs.

```{r}
mu <- list(5, 10, -3)
sigma <- list(1, 5, 10)
seq_along(mu) %>% 
  map(~rnorm(5, mu[[.]], sigma[[.]])) %>% 
  str()
```

But this looks bad. We can use `map2()` or even better `pmap()` that is basically `zip()` on Python.

```{r}
map2(mu, sigma, rnorm, n = 5) %>% str()
```

```{r}
n <- list(1, 3, 5)
args1 <- list(n, mu, sigma)
args1 %>%
  pmap(rnorm) %>% 
  str()
```

## 21.7.1 Invoking different functions

We can use `invoke_map()` as some kind of `eval()` in python.

```{r}
f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)

invoke_map(f, param, n = 5) %>% str()
```

# 21.8 Walk

`walk()` is an alternative to `map()` that can be used when we want the side effect (print, plot, saving) of the funcion instead of the return value

```{r}
x <- list(1, "a", 3)

x %>% 
  walk(print)
```

Another example saving a couple of images using the general `pwalk()`:

```{r}
library(ggplot2)
plots <- mtcars %>% 
  split(.$cyl) %>% 
  map(~ggplot(., aes(mpg, wt)) + geom_point())
paths <- stringr::str_c(names(plots), ".pdf")

pwalk(list(paths, plots), ggsave, path = "output")
```

# 21.9 Other patterns of for loops

## 21.9.1 Predicate functions

keep() and discard() keep elements of the input where the predicate is TRUE or FALSE respectively:

```{r}
iris %>% 
  keep(is.factor) %>% 
  str()

iris %>% 
  discard(is.factor) %>% 
  str()
```

some() and every() determine if the predicate is true for any or for all of the elements.

```{r}
x <- list(1:5, letters, list(10))

x %>% 
  some(is_character)

x %>% 
  every(is_character)

x %>% 
  every(is_vector)
```

detect() finds the first element where the predicate is true; detect_index() returns its position.

```{r}
x <- sample(10)
x

x %>% 
  detect(~ . > 5)

x %>% 
  detect_index(~ . > 5)
```

head_while() and tail_while() take elements from the start or end of a vector while a predicate is true:

```{r}
x

x %>% 
  head_while(~ . > 5)

x %>% 
  tail_while(~ . > 5)
```

## 21.9.2 Reduce and accumulate

This is used when you have multiple items (more useful if its more than 2) and want to reduce them to a singleton.

The reduce function takes a “binary” function (i.e. a function with two primary inputs), and applies it repeatedly to a list until there is only a single element left.

```{r}
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)

dfs %>% reduce(full_join)
```

```{r}
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)

vs %>% reduce(intersect)
```

`accumulate()` is similar but it keeps all the interim results. You could use it to implement a cumulative sum:

```{r}
x <- sample(10)
x

x %>% accumulate(`+`)
```

## 21.9.3 Exercises

1. Implement your own version of every() using a for loop. Compare it with purrr::every(). What does purrr’s version do that your version doesn’t?

```{r}
purrr::every
```

`purrr::every()` returns `NA` if any item is `NA`.

2. Create an enhanced col_sum() that applies a summary function to every numeric column in a data frame.

```{r}
col_sum2 <- function(df, f, ...) {
  map(keep(df, is.numeric), f, ...)
}

col_sum2(iris, mean)
```

3. A possible base R equivalent of col_sum() is:

```{r}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
```

But it has a number of bugs as illustrated with the following inputs:

```{r}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```

What causes the bugs?

The problem is that `sapply` does not always return numeric vectors.
If no columns are selected, instead of returning an empty numeric vector, it returns an empty list.
This causes an error since we can't use a list with `[`.

















