---
title: "24-model-building"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(modelr)
options(na.action = na.warn)

library(nycflights13)
library(lubridate)
```

# 24.2 Why are low quality diamonds more expensive?

We saw this in previous chapters:

```{r}
ggplot(diamonds, aes(cut, price)) + geom_boxplot()
ggplot(diamonds, aes(color, price)) + geom_boxplot()
ggplot(diamonds, aes(clarity, price)) + geom_boxplot()
```

`Fair` cut is the worst. Color `J` is the worst, slightly yellow. Clarity `I1` is the worst, inclusions visible. All this characteristics on their seams to be higher prices.

## 24.2.1 Price and carat

There is one confounding variable (is a variable that influences both the dependent variable and independent variable causing a spurious association). The weight (`carat`) of the diamond is the single most important factor for determining the price of the diamond, and lower quality diamonds tend to be larger.

1. Focus on diamonds smaller than 2.5 carats (99.7% of the data)
2. Log-transform the carat and price variables.

```{r}
diamonds2 <- diamonds %>% 
  filter(carat <= 2.5) %>% 
  mutate(lprice = log2(price), lcarat = log2(carat))
```

```{r}
ggplot(diamonds2, aes(lcarat, lprice)) + 
  geom_hex(bins = 50)
```

```{r}
mod_diamond <- lm(lprice ~ lcarat, data = diamonds2)

grid <- diamonds2 %>% 
  data_grid(carat = seq_range(carat, 20)) %>%   # sequence of 20 between the min and max of carat
  mutate(lcarat = log2(carat)) %>% 
  add_predictions(mod_diamond, "lprice") %>% 
  mutate(price = 2 ^ lprice)

ggplot(diamonds2, aes(carat, price)) + 
  geom_hex(bins = 50) + 
  geom_line(data = grid, colour = "red", size = 1)
```

That tells us something interesting about our data. If we believe our model, then the large diamonds are much cheaper than expected.

The residual shows that we see the right pattern:

```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond, "lresid")

ggplot(diamonds2, aes(lcarat, lresid)) + 
  geom_hex(bins = 50)
```

If we re-do the motivating plots with the residuals instead of `price`:

```{r}
ggplot(diamonds2, aes(cut, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(color, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(clarity, lresid)) + geom_boxplot()
```

Now we see the relationship we expect: as the quality of the diamond increases, so too does its relative price. To interpret the y axis, we need to think about what the residuals are telling us, and what scale they are on. A residual of `-1` indicates that `lprice` was 1 unit lower than a prediction based solely on its weight. A residual of `2^(-1) = 1/2`, points with a value of -1 are half the expected price, and residuals with value 1 are twice the predicted price.

## 24.2.2 A more complicated model

We can include color, cut, and clarity into the model so that we also make explicit the effect of these three categorical variables:

```{r}
mod_diamond2 <- lm(lprice ~ lcarat + color + cut + clarity, data = diamonds2)
```

```{r}
grid <- diamonds2 %>% 
  data_grid(cut, .model = mod_diamond2) %>% 
  add_predictions(mod_diamond2)
grid
```

```{r}
ggplot(grid, aes(cut, pred)) + 
  geom_point()
```

```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond2, "lresid2")

ggplot(diamonds2, aes(lcarat, lresid2)) + 
  geom_hex(bins = 50)
```

This plot indicates that there are some diamonds with quite large residuals - remember a residual of 2 indicates that the diamond is 4x the price that we expected

## 24.2.3 Exercises

1. In the plot of lcarat vs. lprice, there are some bright vertical strips. What do they represent?

```{r}
ggplot(diamonds2, aes(lcarat, lprice)) + 
  geom_hex(bins = 50)
```

The distribution of diamonds has more diamonds at round (human friendly) numbers.

2. If log(price) = a_0 + a_1 * log(carat), what does that say about the relationship between price and carat?

A `1%` increase in carat is associated with an `a_1%` increase in price.

3. Extract the diamonds that have very high and very low residuals. Is there anything unusual about these diamonds? Are they particularly bad or good, or do you think these are pricing errors?

4. Does the final model, mod_diamond2, do a good job of predicting diamond prices? Would you trust it to tell you how much to spend if you were buying a diamond?

```{r}
diamonds2 %>%
  add_predictions(mod_diamond2) %>%
  add_residuals(mod_diamond2) %>%
  summarise(sq_err = sqrt(mean(resid ^ 2)),
            abs_err = mean(abs(resid)),
            p975_err = quantile(resid, 0.975),
            p025_err = quantile(resid, 0.025))
```

Looks good to me XD.

# 24.3 What affects the number of daily flights?

Let’s get started by counting the number of flights per day and visualising it with ggplot2.

```{r}
daily <- flights %>% 
  mutate(date = make_date(year, month, day)) %>% 
  group_by(date) %>% 
  summarise(n = n())
daily

ggplot(daily, aes(date, n)) + 
  geom_line()
```

## 24.3.1 Day of week

Understanding the long-term trend is challenging because there’s a very strong day-of-week effect that dominates the subtler patterns. Let’s start by looking at the distribution of flight numbers by day-of-week:

```{r}
daily <- daily %>% 
  mutate(wday = wday(date, label = TRUE))
ggplot(daily, aes(wday, n)) + 
  geom_boxplot()
```

There are fewer flights on weekends because most travel is for business. The effect is particularly pronounced on Saturday: you might sometimes leave on Sunday for a Monday morning meeting, but it’s very rare that you’d leave on Saturday as you’d much rather be at home with your family.

One way to remove this strong pattern is to use a model. First, we fit the model, and display its predictions overlaid on the original data:

```{r}
mod <- lm(n ~ wday, data = daily)

grid <- daily %>% 
  data_grid(wday) %>% 
  add_predictions(mod, "n")

ggplot(daily, aes(wday, n)) + 
  geom_boxplot() +
  geom_point(data = grid, colour = "red", size = 4)
```

Next we compute and visualise the residuals:

```{r}
daily <- daily %>% 
  add_residuals(mod)

daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

This is quite interesting as the residuals seem a little "stable" until June, after that the changes in the y-axis are higher, now we are seeing the deviation from the expected number of flights, given the day of week. This plot is useful because now that we’ve removed much of the large day-of-week effect, we can see some of the subtler patterns that remain:

1. Our model seems to fail starting in June: you can still see a strong regular pattern that our model hasn’t captured. Drawing a plot with one line for each day of the week makes the cause easier to see. Our model fails to accurately predict the number of flights on Saturday: during summer there are more flights than we expect, and during Fall there are fewer. We’ll see how we can do better to capture this pattern in the next section.

```{r}
ggplot(daily, aes(date, resid, colour = wday)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

2. There are some days with far fewer flights than expected. If you’re familiar with American public holidays, you might spot New Year’s day, July 4th, Thanksgiving and Christmas. There are some others that don’t seem to correspond to public holidays:

```{r}
daily %>% 
  filter(resid < -100)
```

3. There seems to be some smoother long term trend over the course of a year. We can highlight that trend with geom_smooth(). There are fewer flights in January (and December), and more in summer (May-Sep). We can’t do much with this pattern quantitatively, because we only have a single year of data. But we can use our domain knowledge to brainstorm potential explanations.

```{r}
daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line(colour = "grey50") + 
  geom_smooth(se = FALSE, span = 0.20)
```

## 24.3.2 Seasonal Saturday effect

Plot just the Saturdays.

```{r}
daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n)) + 
    geom_point() + 
    geom_line() +
    scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

Lets create a “term” variable that roughly captures the three school terms, and check our work with a plot:

```{r}
term <- function(date) {
  cut(date, 
    breaks = ymd(20130101, 20130605, 20130825, 20140101),
    labels = c("spring", "summer", "fall") 
  )
}

daily <- daily %>% 
  mutate(term = term(date)) 

daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n, colour = term)) +
  geom_point(alpha = 1/3) + 
  geom_line() +
  scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

This improves our model, but not as much as we might hope:

```{r}
mod1 <- lm(n ~ wday, data = daily)
mod2 <- lm(n ~ wday * term, data = daily)

daily %>% 
  gather_residuals(without_term = mod1, with_term = mod2) %>% 
  ggplot(aes(date, resid, colour = model)) +
    geom_line(alpha = 0.75)
```

We can alleviate this problem by using a model that is robust to the effect of outliers: MASS::rlm(). This greatly reduces the impact of the outliers on our estimates, and gives a model that does a good job of removing the day of week pattern:

```{r}
mod3 <- MASS::rlm(n ~ wday * term, data = daily)

daily %>% 
  add_residuals(mod3, "resid") %>% 
  ggplot(aes(date, resid)) + 
  geom_hline(yintercept = 0, size = 2, colour = "white") + 
  geom_line()
```

## 24.3.3 Computed variables

We should be bundling the new variables as functions so we can reuse it.

```{r}
compute_vars <- function(data) {
  data %>% 
    mutate(
      term = term(date), 
      wday = wday(date, label = TRUE)
    )
}
```

Another option is to put the transformations directly in the model formula:

```{r}
wday2 <- function(x) wday(x, label = TRUE)
mod3 <- lm(n ~ wday2(date) * term(date), data = daily)
```

## 24.3.4 Time of year: an alternative approach

Before we used the domain knowledge (US School terms). An alternative to using our knowledge explicitly in the model is to give the data more room to speak. We could use a more flexible model and allow that to capture the pattern we’re interested in.

```{r}
library(splines)
mod <- MASS::rlm(n ~ wday * ns(date, 5), data = daily)

daily %>% 
  data_grid(wday, date = seq_range(date, n = 13)) %>% 
  add_predictions(mod) %>% 
  ggplot(aes(date, pred, colour = wday)) + 
    geom_line() +
    geom_point()
```

## 24.3.5 Exercises

1. Use your Google sleuthing skills to brainstorm why there were fewer than expected flights on Jan 20, May 26, and Sep 1. (Hint: they all have the same explanation.) How would these days generalise to another year?

2. What do the three days with high positive residuals represent? How would these days generalise to another year?

```{r}
daily %>% 
  top_n(3, resid)
#> # A tibble: 3 x 5
#>   date           n wday  resid term 
#>   <date>     <int> <ord> <dbl> <fct>
#> 1 2013-11-30   857 Sat   112.  fall 
#> 2 2013-12-01   987 Sun    95.5 fall 
#> 3 2013-12-28   814 Sat    69.4 fall

```

3. Create a new variable that splits the wday variable into terms, but only for Saturdays, i.e. it should have Thurs, Fri, but Sat-summer, Sat-spring, Sat-fall. How does this model compare with the model with every combination of wday and term?

4. Create a new wday variable that combines the day of week, term (for Saturdays), and public holidays. What do the residuals of that model look like?

5. What happens if you fit a day of week effect that varies by month (i.e. n ~ wday * month)? Why is this not very helpful?

6. What would you expect the model n ~ wday + ns(date, 5) to look like? Knowing what you know about the data, why would you expect it to be not particularly effective?

7. We hypothesised that people leaving on Sundays are more likely to be business travellers who need to be somewhere on Monday. Explore that hypothesis by seeing how it breaks down based on distance and time: if it’s true, you’d expect to see more Sunday evening flights to places that are far away.

8. It’s a little frustrating that Sunday and Saturday are on separate ends of the plot. Write a small function to set the levels of the factor so that the week starts on Monday.
