---
title: "25-many-models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(modelr)
library(tidyverse)
```

# 25.2 gapminder

```{r}
library(gapminder)
gapminder
```

The question we want to answer is: “How does life expectancy (lifeExp) change over time (year) for each country (country)?”.

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)
```

Overall, it looks like life expectancy has been steadily improving. However, if you look closely, you might notice some countries that don’t follow this pattern. How can we make those countries easier to see?

One way is to use the same approach as in the last chapter: there’s a strong signal (overall linear growth) that makes it hard to see subtler trends. We’ll tease these factors apart by fitting a model with a linear trend. The model captures steady growth over time, and the residuals will show what’s left.

Let's do it for one country:

```{r}
nz <- filter(gapminder, country == "New Zealand")
nz %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_line() + 
  ggtitle("Full data = ")

```

```{r}
nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>% 
  add_predictions(nz_mod) %>%
  ggplot(aes(year, pred)) + 
  geom_line() + 
  ggtitle("Linear trend + ")

nz %>% 
  add_residuals(nz_mod) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Remaining pattern")
```

## 25.2.1 Nested data

How can we easily fit the previous model to every country?

To do that, we need a new data structure: the nested data frame. To create a nested data frame we start with a grouped data frame, and “nest” it:

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()

by_country
```

This creates a data frame that has one row per group (per country), and a rather unusual column: data. data is a list of data frames (or tibbles, to be precise)

We can get one item from that df, in this case Afganistan.

```{r}
by_country$data[[1]]
```

## 25.2.2 List-columns

Now that we have our nested data frame, we’re in a good position to fit some models. We have a model-fitting function:

```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
```

```{r}
by_country <- by_country %>% 
  mutate(model = map(data, country_model))
by_country
```

Keeping the data and models in a single df as a new column its very nice because

## 25.2.3 Unnesting

Let's start by calculating the residuals by adding them to a new column.

```{r}
by_country <- by_country %>% 
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country
```

But how you can plot a list of data frames? Instead of struggling to answer that question, let’s turn the list of data frames back into a regular data frame. Previously we used nest() to turn a regular data frame into an nested data frame, and now we do the opposite with unnest():

```{r}
resids <- unnest(by_country, resids)
resids
```

```{r}
resids %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group = country), alpha = 1 / 3) + 
    geom_smooth(se = FALSE)
```

```{r}
resids %>% 
  ggplot(aes(year, resid, group = country)) +
    geom_line(alpha = 1 / 3) + 
    facet_wrap(~continent)
```

There is some missing patters specially on Afreica and Asia.

## 25.2.4 Model quality

Instead of looking at the residuals from the model, we could look at some general measurements of model quality.

```{r}
broom::glance(nz_mod)
```

We can do this on the nested dataframe and using `.drop = TRUE` since we are unnesting just by the `glance` column.

```{r}
glance <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE)
glance
```

Which models do not fit well?

```{r}
glance %>% 
  arrange(r.squared)
```

The worst models appear to be in Africa.

```{r}
glance %>% 
  ggplot(aes(continent, r.squared)) + 
    geom_jitter(width = 0.5)
```

Pull and plot the contries that do not fit well:

```{r}
bad_fit <- filter(glance, r.squared < 0.25)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot(aes(year, lifeExp, colour = country)) +
    geom_line()
```

## 25.2.5 Exercises

1. A linear trend seems to be slightly too simple for the overall trend. Can you do better with a quadratic polynomial? How can you interpret the coefficients of the quadratic? (Hint you might want to transform year so that it has mean zero.)

```{r}
country_model <- function(df) {
  lm(lifeExp ~ poly(year - median(year), 2), data = df)
}

by_country <- gapminder %>%
  group_by(country, continent) %>%
  nest()

by_country <- by_country %>%
  mutate(model = map(data, country_model))
```

```{r}
by_country <- by_country %>%
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country
```

```{r}
unnest(by_country, resids) %>%
ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1 / 3) +
  geom_smooth(se = FALSE)
```

```{r}
by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance, .drop = TRUE) %>%
  ggplot(aes(continent, r.squared)) +
  geom_jitter(width = 0.5)
```

2. Explore other methods for visualising the distribution of R^2 per continent. You might want to try the ggbeeswarm package, which provides similar methods for avoiding overlaps as jitter, but uses deterministic methods.

```{r}
library("ggbeeswarm")
by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance, .drop = TRUE) %>%
  ggplot(aes(continent, r.squared)) +
  geom_beeswarm()
```

3. To create the last plot (showing the data for the countries with the worst model fits), we needed two steps: we created a data frame with one row per country and then semi-joined it to the original dataset. It’s possible to avoid this join if we use unnest() instead of unnest(.drop = TRUE). How?

```{r}
gapminder %>% 
  group_by(country, continent) %>% 
  nest() %>% 
  mutate(model = map(data, ~ lm(lifeExp ~ year, .))) %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance) %>%
  unnest(data) %>%
  filter(r.squared < 0.25) %>%
  ggplot(aes(year, lifeExp)) + 
    geom_line(aes(color = country))
```

# 25.3 List-columns

R default data.frame doesn't allow to create list-columns easily, we have to use `I()`.

```{r}
data.frame(x = list(1:3, 3:5))

data.frame(
  x = I(list(1:3, 3:5)), 
  y = c("1, 2", "3, 4, 5")
)
```

It's easier with tibbles:

```{r}
tibble(
  x = list(1:3, 3:5), 
  y = c("1, 2", "3, 4, 5")
)
```

List-columns are often most useful as intermediate data structure. They’re hard to work with directly, because most R functions work with atomic vectors or data frames, but the advantage of keeping related items together in a data frame is worth a little hassle.

Generally there are three parts of an effective list-column pipeline:

1. You create the list-column using one of nest(), summarise() + list(), or mutate() + a map function, as described in Creating list-columns.
2. You create other intermediate list-columns by transforming existing list columns with map(), map2() or pmap(). For example, in the case study above, we created a list-column of models by transforming a list-column of data frames.
3. You simplify the list-column back down to a data frame or atomic vector, as described in Simplifying list-columns.

# 25.4 Creating list-columns

Typically, you won’t create list-columns with tibble(). Instead, you’ll create them from regular columns, using one of three methods:

1. With `tidyr::nest()` to convert a grouped data frame into a nested data frame where you have list-column of data frames.
2. With `mutate()` and vectorised functions that return a list.
3. With `summarise()` and summary functions that return multiple results.

Generally, when creating list-columns, you should make sure they’re homogeneous: each element should contain the same type of thing. There are no checks to make sure this is true, but if you use purrr and remember what you’ve learned about type-stable functions, you should find it happens naturally.

## 25.4.1 With nesting

We could do it with a grouped data frame.

```{r}
gapminder %>% 
  group_by(country, continent) %>% 
  nest()
```

Or an ungrouped one:

```{r}
gapminder %>% 
  nest(year:gdpPercap)
```

## 25.4.2 From vectorised functions

Some useful functions take an atomic vector and return a list. For example, in strings you learned about stringr::str_split() which takes a character vector and returns a list of character vectors. If you use that inside mutate, you’ll get a list-column:

```{r}
df <- tribble(
  ~x1,
  "a,b,c", 
  "d,e,f,g"
) 

df %>% 
  mutate(x2 = stringr::str_split(x1, ","))
```

`unnest()` knows how to handle this vectors.

```{r}
df %>% 
  mutate(x2 = stringr::str_split(x1, ",")) %>% 
  unnest()
```

Another example of this pattern is using the map(), map2(), pmap() from purrr.

```{r}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)

sim %>%
  mutate(sims = invoke_map(f, params, n = 10))
```

## 25.4.3 From multivalued summaries

One restriction of summarise() is that it only works with summary functions that return a single value. That means that you can’t use it with functions like quantile() that return a vector of arbitrary length:

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = quantile(mpg))
```

You can however, wrap the result in a list! This obeys the contract of summarise(), because each summary is now a list (a vector) of length 1.

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg)))
```

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg))) %>% 
  unnest()
```

## 25.4.4 From a named list

Data frames with list-columns provide a solution to a common problem: what do you do if you want to iterate over both the contents of a list and its elements?

```{r}
x <- list(
  a = 1:5,
  b = 3:4, 
  c = 5:6
) 

df <- enframe(x)
df
```

Now if you want to iterate over names and values in parallel, you can use map2():

```{r}
df %>% 
  mutate(
    smry = map2_chr(name, value, ~ stringr::str_c(.x, ": ", .y[1]))
  )
```

## 25.4.5 Exercises

1. List all the functions that you can think of that take a atomic vector and return a list.

`stringr` functions.

2. Brainstorm useful summary functions that, like quantile(), return multiple values.

3. What’s missing in the following data frame? How does quantile() return that missing piece? Why isn’t that helpful here?

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg))) %>% 
  unnest()
```

The particular quantiles of the values are missing, e.g. `0%`, `25%`, `50%`, `75%`, `100%`. `quantile()` returns these in the names of the vector.

The right one is as mentioned in the book:

```{r}
probs <- c(0.01, 0.25, 0.5, 0.75, 0.99)
mtcars %>% 
  group_by(cyl) %>% 
  summarise(p = list(probs), q = list(quantile(mpg, probs))) %>% 
  unnest()
```

4. What does this code do? Why might might it be useful?

```{r}
?funs
```


```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise_each(funs(list))
```

Creates a df each row is a value of `cyl`  and each column is a vector of all the column values for that row (value of cyl).

# 25.5 Simplifying list-columns

To apply the techniques of data manipulation and visualisation you’ve learned in this book, you’ll need to simplify the list-column back to a regular column (an atomic vector), or set of columns. The technique you’ll use to collapse back down to a simpler structure depends on whether you want a single value per element, or multiple values:

1. If you want a single value, use mutate() with map_lgl(), map_int(), map_dbl(), and map_chr() to create an atomic vector.
2. If you want many values, use unnest() to convert list-columns back to regular columns, repeating the rows as many times as necessary.

## 25.5.1 List to vector

```{r}
df <- tribble(
  ~x,
  letters[1:5],
  1:3,
  runif(5)
)
df

df %>% mutate(
  type = map_chr(x, typeof),
  length = map_int(x, length)
)
```

```{r}
df <- tribble(
  ~x,
  list(a = 1, b = 2),
  list(a = 2, c = 4)
)
df

df %>% mutate(
  a = map_dbl(x, "a"),
  b = map_dbl(x, "b", .null = NA_real_)
)
```

## 25.5.2 Unnesting

`unnest()` works by repeating the regular columns once for each element of the list-column.

```{r}
df <- tibble(x = 1:2, y = list(1:4, 1))
df

df %>%
  unnest(y)
```

This means that you can’t simultaneously unnest two columns that contain different number of elements. You can if they have the same number of elements.

```{r}
# Doesn't work because y and z have different number of elements
df2 <- tribble(
  ~x, ~y,           ~z,
   1, "a",         1:2,  
   2, c("b", "c"),   3
)
df2

df2 %>% unnest(y, z)
```

```{r}
# Ok, because y and z have the same number of elements in every row
df1 <- tribble(
  ~x, ~y,           ~z,
   1, c("a", "b"), 1:2,
   2, "c",           3
)
df1

df1 %>% unnest(y, z)
```

## 25.5.3 Exercises

1. Why might the lengths() function be useful for creating atomic vector columns from list-columns?

It could be useful for testing whether all elements in a list-column are the same length.

2. List the most common types of vector found in a data frame. What makes lists different?

The common types of vectors in data frames are:

-   `logical`
-   `numeric`
-   `integer`
-   `character`
-   `factor`

All of the common types of vectors in data frames are atomic. Lists are not atomic (they can contain other lists and other vectors).

# 25.6 Making tidy data with broom

The broom package provides three general tools for turning models into tidy data frames:

1. `broom::glance(model)` returns a row for each model. Each column gives a model summary: either a measure of model quality, or complexity, or a combination of the two.
2. `broom::tidy(model)` returns a row for each coefficient in the model. Each column gives information about the estimate or its variability.
3. `broom::augment(model, data)` returns a row for each row in data, adding extra values like residuals, and influence statistics.
