---
title: "29-rmarkdown-formats"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 29.3 Documents

We can have multiple outputs for one document, each one with different options:

```
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
```

# 29.4 Notebooks

Notebooks are a special type of `html_document`. The main difference is that a notebook is used for collaboration and generates a `.nb.html` file that also includes all the code. R Studio can open this file and recreate the `.rmd` that created it.

We can use `github_document` to generate a git friendly version of the notebook.

```
output:
  html_notebook: default
  github_document: default
```

# 29.5 Presentations

R Markdown comes with three presentation formats built-in:

1. `ioslides_presentation` - HTML presentation with ioslides
2. `slidy_presentation` - HTML presentation with W3C Slidy
3. `beamer_presentation` - PDF presentation with LaTeX Beamer.

Two other popular formats are provided by packages:

1. `revealjs::revealjs_presentation` - HTML presentation with reveal.js. Requires the revealjs package.
2. `rmdshower`, https://github.com/MangoTheCat/rmdshower, provides a wrapper around the shower, https://github.com/shower/shower, presentation engine

# 29.6 Dashboards

We can use Flexdashboards as an output. See the example: `29-rmarkdown-formats-flex-dashboard.rmd`.

```{r}
# install.packages("flexdashboard")
# install.packages("DT")
```

# 29.7 Interactivity

## 29.7.1 htmlwidgets

Some libraries provide interactive html widgets such as `leaflet`.

```{r}
# install.packages("leaflet")
```

```{r}
library(leaflet)
leaflet() %>%
  setView(174.764, -36.877, zoom = 16) %>% 
  addTiles() %>%
  addMarkers(174.764, -36.877, popup = "Maungawhau") 
```

There are many packages that provide htmlwidgets, including:

- dygraphs, http://rstudio.github.io/dygraphs/, for interactive time series visualisations.
- DT, http://rstudio.github.io/DT/, for interactive tables.
- threejs, https://github.com/bwlewis/rthreejs for interactive 3d plots.
- DiagrammeR, http://rich-iannone.github.io/DiagrammeR/ for diagrams (like flow charts and simple node-link diagrams).

## 29.7.2 Shiny

htmlwidgets provide only client-side interactivity via javascript. If we want to use R to do more computation we can use Shiny for this. We can set a rmarkdown document to use shiny with:

```
title: "Shiny Web App"
output: html_document
runtime: shiny
```

```{r}
library(shiny)

textInput("name", "What is your name?")
numericInput("age", "How old are you?", NA, min = 0, max = 150)
```



